---
title: "BB512/BB612 - Week II"
output: pdf_document
---

```{r}
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(Biobase))
suppressPackageStartupMessages(library(preprocessCore))
```

# Data

We'll explore the `bodymap` expression dataset.

```{r data}
con <- url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)

bm <- bodymap.eset

pdata <- pData(bm) # phenotype data
edata <- exprs(bm) # expression data
fdata <- fData(bm) # features data
```

# Phenotype data

We'll first explore the phenotype data:

```{r p_eda1}
head(pdata,3)
```

```{r p_eda2}
table(pdata$gender)

for_plot_df <- as.data.frame(table(pdata$gender))
head(for_plot_df)
g <- ggbarplot(for_plot_df, x = "Var1", y = "Freq")
g

g2 <- ggbarplot(for_plot_df, x = "Var1", y = "Freq", color = "Var1")
g2

g3 <- ggbarplot(for_plot_df, x = "Var1", y = "Freq", fill = "Var1")
g3
```

```{r p_eda3}
table(pdata$gender, pdata$race)

for_plot_df2 <- as.data.frame(table(pdata$gender, pdata$race))
head(for_plot_df2)

g4 <- ggbarplot(for_plot_df2, "Var1", "Freq", fill = "Var2")
g4
```

```{r p_eda4}
summary(pdata$age)

ggboxplot(pdata, "gender", "age")

g5 <- ggboxplot(pdata, "gender", "age", color = "gender", add = "jitter")
g5


g6 <- ggviolin(pdata, "gender", "age", color = "gender", add = c("boxplot", "jitter"))
g6
```

# Expression data

Look at overall distributions:

```{r norm1}
boxplot(log2(edata + 1))
```

alternatives:

```{r norm2}
# sample by sample histograms
par(mfrow = c(1, 2))
hist(log2(edata[, 1] + 1), col = 2)
hist(log2(edata[, 2] + 1), col = 2)
par(mfrow = c(1, 1))

### or with densinty plots
plot(density(log2(edata[, 1] + 1)), col = "red")
for (i in 2:ncol(edata)) {
    lines(density(log2(edata[, 2] + 1)), col = "red")
}
```

We can remove rows that are mostly zero and notice any differences in the distributions across samples.

```{r norm3}
filt_edata <- edata[rowMeans(edata) > 1, ]
boxplot(log2(filt_edata + 1))
```

A common type of plot for genomics data is a heatmap. They are usually used for visualizing matrices. For example we can look at all genes with an average number of counts greater than 10000:

```{r norm4}
ematrix <- as.matrix(filt_edata)[rowMeans(filt_edata) > 10000, ]
heatmap(ematrix)
```

Quantile normalization:

```{r norm5}
lt_filt_edata <- log2(filt_edata + 1)
norm_edata <- normalize.quantiles(lt_filt_edata)
colnames(norm_edata) <- colnames(lt_filt_edata)
rownames(norm_edata) <- rownames(lt_filt_edata)
boxplot(norm_edata)
```