---
title: "BB512/BB612 - Week III"
output: pdf_document
---

```{r}
suppressPackageStartupMessages(library(Biobase))
suppressPackageStartupMessages(library(dendextend))
suppressPackageStartupMessages(library(factoextra))
```

# Clustering

## Data

Once again, we'll use the `bodymap` expression dataset.

```{r cl_data}
con <- url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file = con)
close(con)

bm <- bodymap.eset

pdata <- pData(bm) # phenotype data
edata <- exprs(bm) # expression data
fdata <- fData(bm) # features data
```

```{r proc}
edata <- edata[rowMeans(edata) > 5000, ]
edata <- log2(edata + 1)
```

## Distances

```{r dist}
# By default calculates the euclidean distance between rows
dist1 = dist(t(edata))

## Look at distance matrix
colramp <- colorRampPalette(c("blue", "white", "red"))(9)
heatmap(as.matrix(dist1), col = colramp, Colv = NA, Rowv = NA)
```

## Hierarchical clustering

Here we use the distance we previously calculated to perform a hierarchical clustering and plot the dendrogram:

```{r hclu1}
hclust1 <- hclust(dist1)
plot(hclust1)
plot(hclust1, hang = -1)
```

We can also color the dendrogram either into a fixed number of groups:

```{r hclu2}
dend <- as.dendrogram(hclust1)
dend <- color_labels(hclust1,4,col=1:4)
plot(dend)
```

Or you can color them directly:

```{r hclu3}
labels_colors(dend) <- as.numeric(pdata$tissue.type[match(labels(dend), pdata$sample.id)])
plot(dend)
```

## K-means clustering

Now we can perform k-means clustering. By default, the rows are clustered. You can either input the cluster means (often unknown) or the number of clusters:

```{r kmeans1}
set.seed(123)
kmeans1 <- kmeans(t(edata), centers = 3)
names(kmeans1)
```

Now we can look at the cluster centers:

```{r kmeans2}
matplot(t(kmeans1$centers), col = 1:3, type = "l", lwd = 3)
```

We can observe which points are assigned to which cluster

```{r kmeans3}
kmeans1$cluster
```

```{r km_hm}
heatmap(as.matrix(edata)[, order(kmeans1$cluster)], col = colramp, Colv = NA, Rowv = NA)
```

## Biclustering

```{r bicl}
heatmap(as.matrix(edata))
```

# Dimensionality Reduction

See [this article](http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/) for more detail.

## Data

```{r dr_data}
data(decathlon2)
decathlon2.active <- decathlon2[1:23, 1:10]
head(decathlon2.active)
```


## Compute PCA

```{r pca1}
res.pca <- prcomp(decathlon2.active, scale = TRUE)
```

Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component:

```{r pca2}
fviz_eig(res.pca)
```

## Plots

Graph of individuals. Individuals with a similar profile are grouped together.

```{r pca3}
fviz_pca_ind(res.pca,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

Graph of variables. Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph.

```{r pca4}
fviz_pca_var(res.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

Biplot:

```{r biplot}
fviz_pca_biplot(res.pca, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

## Access to PCA results

```{r pca_res}
# Eigenvalues
eig.val <- get_eigenvalue(res.pca)
eig.val
  
# Results for Variables
res.var <- get_pca_var(res.pca)
res.var$coord          # Coordinates
res.var$contrib        # Contributions to the PCs
res.var$cos2           # Quality of representation 
# Results for individuals
res.ind <- get_pca_ind(res.pca)
res.ind$coord          # Coordinates
res.ind$contrib        # Contributions to the PCs
res.ind$cos2           # Quality of representation 
```
